---
title: "01 -- Data and the two measures"
subtitle: "FIP_Pyth_WinRate and FIP_Pyth_WL"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
library(tidyverse)
DATA <- "."
OUT  <- "."
```

## Constants

`GAMMA` is the Pythagorean exponent, `UR` the league unearned-run rate. Both are fixed
across every script so that nothing downstream can silently re-anchor the measure.

```{r constants}
GAMMA <- 1.83
UR    <- 0.31
```

## Reading the Baseball-Reference exports

One row per pitcher-team stint. Pitchers who changed teams also get a season-total row
whose `Team` reads `2TM`, `3TM` and so on; those rows are the season totals and are kept
separately from the per-team stints.

```{r read}
read_season <- function(file, season) {
  readr::read_csv(file.path(DATA, file), show_col_types = FALSE) %>%
    rename(pid = `Player-additional`, name = Player, team = Team, lg = Lg) %>%
    mutate(name   = str_remove_all(name, "[*#]") %>% str_trim(),
           season = season,
           multi  = str_detect(team, "^\\d+TM$"))
}

raw <- bind_rows(
  read_season("pitching_2023.txt",               2023),
  read_season("Pitching_2024.txt",               2024),
  read_season("Pitching_2025.txt",               2025),
  read_season("Pitching_2026_through_sept_1.txt", 2026))
```

### Innings are base-three

`183.2` means 183 and two-thirds, not 183.2. Reading it as a decimal understates league
innings by tens of frames and is the single most common parsing error in this data.

```{r innings}
ip_true <- function(x) { w <- floor(x + 1e-9); w + round((x - w) * 10) / 3 }
raw <- raw %>% mutate(IP = ip_true(IP), G_opp = IP / 9)
```

## The win-neutral anchor

League-average FIP does **not** calibrate. The win function is convex above an inflection
near FIP 1.72, so by Jensen's inequality the innings-weighted mean of the mapped win rates
exceeds .500 even when the mean FIP is league average. We therefore solve for the anchor
`F*` at which the population does balance.

```{r anchor}
wp <- function(x, ref) {
  num <- (ref + UR)^GAMMA
  num / (num + pmax(x + UR, 1e-6)^GAMMA)
}

season_totals <- raw %>%
  group_by(season, pid) %>%
  slice(if (any(multi)) which(multi)[1] else 1) %>%
  ungroup() %>%
  filter(!is.na(FIP), !is.na(ERA), IP > 0)

anchors <- season_totals %>%
  group_by(season) %>%
  summarise(
    lg_FIP = weighted.mean(FIP, IP),
    lg_ERA = weighted.mean(ERA, IP),
    FSTAR  = uniroot(function(r) sum(G_opp * wp(FIP, r)) / sum(G_opp) - 0.5,
                     c(0.5, 15), tol = 1e-12)$root,
    ESTAR  = uniroot(function(r) sum(G_opp * wp(ERA, r)) / sum(G_opp) - 0.5,
                     c(0.5, 15), tol = 1e-12)$root,
    naive  = sum(G_opp * wp(FIP, lg_FIP)) / sum(G_opp),
    .groups = "drop") %>%
  mutate(jensen_thousandths = 1000 * (naive - 0.5))
knitr::kable(anchors, digits = 4)
```

`jensen_thousandths` is how far league anchoring misses .500. It runs 11--14 thousandths
on every population, an order of magnitude larger than the same calculation on club
aggregates, because the correction bites where the dispersion is -- across individual
pitchers.

## The two measures

```{r measures}
season_totals <- season_totals %>%
  left_join(anchors %>% select(season, FSTAR, ESTAR), by = "season") %>%
  mutate(fip_rate   = wp(FIP, FSTAR),     # FIP_Pyth_WinRate
         era_rate   = wp(ERA, ESTAR),     # the ERA-anchored comparison
         wedge_rate = era_rate - fip_rate,
         fip_W      = G_opp * fip_rate,   # FIP_Pyth_WL, win side
         fip_L      = G_opp - fip_W)      # FIP_Pyth_WL, loss side
```

`fip_W + fip_L = G_opp` by construction, so a pitcher's record always spans exactly the
opportunities he took.

## Roles, from actual appearance counts

```{r roles}
stints <- raw %>%
  filter(!multi, !is.na(FIP), IP > 0) %>%
  left_join(anchors %>% select(season, FSTAR), by = "season") %>%
  mutate(role = case_when(SV >= 8                 ~ "Closer",
                          G > 0 & GS / G >= 0.60  ~ "Starter",
                          TRUE                    ~ "Relief"),
         fip_rate = wp(FIP, FSTAR),
         fip_W    = G_opp * fip_rate,
         fip_L    = G_opp - fip_W)

stints %>% filter(season == 2026) %>%
  group_by(role) %>%
  summarise(n = n(), IP = sum(IP), FIP = weighted.mean(FIP, IP),
            win_rate = sum(fip_W) / sum(G_opp), .groups = "drop") %>%
  knitr::kable(digits = 3)
```

```{r save}
write_csv(season_totals, file.path(OUT, "pitcher_seasons_full.csv"))
write_csv(stints,        file.path(OUT, "pitcher_stints_full.csv"))
```
