---
title: "02 -- Validation: attribution, persistence, and the map itself"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
library(tidyverse)
DATA <- "../data"
GAMMA <- 1.83; UR <- 0.31
wp <- function(x, ref) { num <- (ref + UR)^GAMMA; num / (num + pmax(x + UR, 1e-6)^GAMMA) }
ip_true <- function(x) { w <- floor(x + 1e-9); w + round((x - w) * 10) / 3 }
seasons <- read_csv(file.path(DATA, "pitcher_seasons_full.csv"), show_col_types = FALSE)
stints  <- read_csv(file.path(DATA, "pitcher_stints_full.csv"),  show_col_types = FALSE)
```

## A. Attribution: which inputs own the variation in winning?

Regressing club wins on staff pitching alone omits the offense, which is 38--42 % of win
variance. That misspecification also flatters ERA, because ERA absorbs fielding and
sequencing. We therefore recover each club's offense by inverting the Pythagorean identity
and decompose run prevention into channels that are actually distinct.

```{r teamprep}
tm <- read_csv(file.path(DATA, "bbref_2026_teams.csv"), show_col_types = FALSE) %>%
  mutate(IP = ip_true(IP_raw), G_opp = IP / 9, games = W + L, wpct = W / games,
         RS_G = RA_G * (wpct / (1 - wpct))^(1 / GAMMA))

agg <- stints %>% filter(season == 2026) %>%
  group_by(team) %>%
  summarise(across(c(IP, H, R, ER, HR, BB, SO, HBP, BF), sum), .groups = "drop")

m <- tm %>% select(team, W, games, RS_G, RA_G, G_opp) %>% left_join(agg, by = "team") %>%
  mutate(bip   = BF - SO - BB - HBP - HR,
         BABIP = (H - HR) / bip,                                   # fielding + fortune
         LOB   = (H + BB + HBP - R) / (H + BB + HBP - 1.4 * HR),   # sequencing
         UER9  = 9 * (R - ER) / IP)                                # fielding errors
```

```{r fipw}
# staff FIP_Pyth_WL wins, scaled to the club's games played
staff_fip <- stints %>% filter(season == 2026) %>%
  group_by(team) %>% summarise(fipw_raw = sum(fip_W), .groups = "drop")
m <- m %>% left_join(staff_fip, by = "team") %>%
  mutate(fipw = fipw_raw * games / G_opp)
```

### The five-input decomposition

```{r decomp}
fit <- lm(W ~ fipw + BABIP + LOB + UER9 + RS_G, data = m)
share <- function(x, y, b) b * cov(x, y) / var(y)
b <- coef(fit)
tibble(component = c("fielding-independent pitching", "balls in play (BABIP)",
                     "sequencing (strand rate)", "fielding errors", "offense"),
       share = c(share(m$fipw, m$W, b[2]), share(m$BABIP, m$W, b[3]),
                 share(m$LOB,  m$W, b[4]), share(m$UER9, m$W, b[5]),
                 share(m$RS_G, m$W, b[6]))) %>%
  mutate(pct = 100 * share) %>% select(-share) %>% knitr::kable(digits = 1)
cat(sprintf("R-squared %.3f; residual %.1f%%\n", summary(fit)$r.squared,
            100 * (1 - summary(fit)$r.squared)))
```

Sequencing, not fielding, is what a run average mostly mistakes for pitching. Errors are
almost nothing. Pitching proper exceeds fielding even charging every ball in play to the
fielders.

## B. Persistence, at the level that matters

This must be run on **individual pitchers**. A club-level version is confounded twice
over: rosters turn over, and a whole staff's fielding averages out in a way that flatters
ERA rather than testing it. Run the club version and sequencing appears *more* persistent
than pitching, which is an artifact of aggregation.

```{r persist}
pairs <- list(c(2023, 2024), c(2024, 2025), c(2025, 2026))
panel <- map_dfr(pairs, function(p) {
  a <- seasons %>% filter(season == p[1]) %>% select(pid, IP, fip_rate, era_rate, wedge_rate)
  b <- seasons %>% filter(season == p[2]) %>% select(pid, IP2 = IP, f2 = fip_rate,
                                                     e2 = era_rate, w2 = wedge_rate)
  inner_join(a, b, by = "pid") %>% mutate(pair = paste0(p[1], "-", p[2]))
})

map_dfr(c(10, 20, 30, 40, 50, 70, 100), function(thr) {
  s <- panel %>% filter(IP >= thr, IP2 >= thr)
  tibble(min_IP = thr, n = nrow(s),
         FIP_rate = cor(s$fip_rate, s$f2),
         ERA_rate = cor(s$era_rate, s$e2),
         wedge    = cor(s$wedge_rate, s$w2))
}) %>% mutate(gap = FIP_rate - ERA_rate) %>% knitr::kable(digits = 3)
```

The FIP-anchored rate repeats about twice as strongly as the ERA-anchored one at every
bar, and the wedge -- exactly what ERA carries and this measure excludes -- barely repeats.

## C. Is the Pythagorean map itself right?

The form and exponent were calibrated on *team* runs. We test them on individual pitchers
using the official won-lost record, which is noisy but close to orthogonal to a pitcher's
own FIP, so its conditional mean given FIP traces the underlying map.

```{r map}
st <- seasons %>% filter(GS / pmax(G, 1) >= 0.80, (W + L) >= 10) %>%
  mutate(awpct = W / (W + L))
curve_fn <- function(fip, gamma, ref) { num <- (ref + UR)^gamma; num / (num + (fip + UR)^gamma) }
fit_map <- nls(awpct ~ curve_fn(FIP, gamma, ref), data = st,
               start = list(gamma = 1.83, ref = 4.1), weights = (W + L))
summary(fit_map)$coefficients %>% knitr::kable(digits = 3)
```

If `gamma` is statistically indistinguishable from 1.83, the functional form inherited
from team run-scoring survives a test it had not previously been given.
