---
title: "03 -- Machine learning and uncertainty"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
library(tidyverse); library(glmnet); library(ranger); library(gbm)
DATA <- "../data"; set.seed(20260901)
seasons <- read_csv(file.path(DATA, "pitcher_seasons_full.csv"), show_col_types = FALSE)
MINIP <- 40
```

## Building the supervised-learning panel

The target is next season's `FIP_Pyth_WinRate`. The thirteen inputs deliberately include
the ERA-anchored rate, so the learner is free to prefer it over ours.

```{r panel}
feat <- function(d) d %>%
  mutate(K9 = 9 * SO / IP, BB9 = 9 * BB / IP, HR9 = 9 * HR / IP, Hit9 = 9 * H / IP,
         KBB = SO / pmax(BB, 1), GS_share = GS / pmax(G, 1), SV_rate = SV / pmax(G, 1),
         BABIP = (H - HR) / pmax(BF - SO - BB - HBP - HR, 1), Workload = IP)

FEATS <- c("fip_rate","era_rate","K9","BB9","HR9","Hit9","KBB","WHIP",
           "BABIP","Workload","Age","GS_share","SV_rate")

make_pair <- function(y1, y2, split) {
  a <- feat(seasons %>% filter(season == y1))
  b <- seasons %>% filter(season == y2) %>% select(pid, y = fip_rate)
  inner_join(a, b, by = "pid") %>% filter(IP >= MINIP) %>% mutate(split = split)
}
d  <- bind_rows(make_pair(2023, 2024, "train"),
                make_pair(2024, 2025, "calib"),
                make_pair(2025, 2026, "test")) %>% drop_na(all_of(c(FEATS, "y")))
tr <- d %>% filter(split == "train"); ca <- d %>% filter(split == "calib")
te <- d %>% filter(split == "test")
cat(sprintf("train %d | calibrate %d | test %d\n", nrow(tr), nrow(ca), nrow(te)))
```

Training uses 2023-24 and 2024-25; 2025-26 is held out entirely, with no season shared.

## Baselines and three learners

```{r models}
trf <- bind_rows(tr, ca)
Xtr <- as.matrix(trf[FEATS]); Xte <- as.matrix(te[FEATS])

r2 <- function(p, y) 1 - sum((y - p)^2) / sum((y - mean(y))^2)
res <- tibble()

res <- bind_rows(res, tibble(model = "Unconditional mean",
  R2 = r2(rep(mean(trf$y), nrow(te)), te$y)))
for (v in c("fip_rate", "era_rate")) {
  f <- lm(reformulate(v, "y"), data = trf)
  res <- bind_rows(res, tibble(model = paste0("AR(1) on ", v),
                               R2 = r2(predict(f, te), te$y)))
}

ridge <- cv.glmnet(Xtr, trf$y, alpha = 0)
res <- bind_rows(res, tibble(model = "Ridge (all 13)",
  R2 = r2(as.numeric(predict(ridge, Xte, s = "lambda.min")), te$y)))

rf <- ranger(y ~ ., data = trf[c(FEATS, "y")], num.trees = 600,
             min.node.size = 5, mtry = 6, importance = "permutation")
res <- bind_rows(res, tibble(model = "Random forest",
  R2 = r2(predict(rf, te)$predictions, te$y)))

gb <- gbm(y ~ ., data = trf[c(FEATS, "y")], distribution = "gaussian",
          n.trees = 400, shrinkage = 0.03, interaction.depth = 3,
          bag.fraction = 0.8, verbose = FALSE)
res <- bind_rows(res, tibble(model = "Gradient boosting",
  R2 = r2(predict(gb, te, n.trees = 400), te$y)))
knitr::kable(res, digits = 3)
```

## Permutation importance on the held-out set

This is the decisive quantity: what a *fitted* model leans on out of sample, not what it
split on during training.

```{r perm}
base <- r2(predict(gb, te, n.trees = 400), te$y)
imp <- map_dfr(FEATS, function(f) {
  drops <- replicate(40, {
    sh <- te; sh[[f]] <- sample(sh[[f]])
    base - r2(predict(gb, sh, n.trees = 400), te$y)
  })
  tibble(feature = f, importance = mean(drops), sd = sd(drops))
}) %>% arrange(desc(importance))
knitr::kable(imp, digits = 4)
```

If `fip_rate` ranks first of thirteen and several times above `era_rate`, the learner has
chosen the fielding-independent rate while free to choose otherwise.

## Conformal prediction intervals

Quantile boosting alone under-covers out of sample. The correction must be calibrated on
a *held-out transition*, not a random slice of the training years, because the shortfall
comes from year-to-year drift a within-year slice cannot see.

```{r conformal}
qfit <- function(a) gbm(y ~ ., data = tr[c(FEATS, "y")], distribution = list(name = "quantile", alpha = a),
                        n.trees = 400, shrinkage = 0.03, interaction.depth = 3,
                        bag.fraction = 0.8, verbose = FALSE)
q10 <- qfit(0.10); q50 <- qfit(0.50); q90 <- qfit(0.90)
pr <- function(m, nd) predict(m, nd, n.trees = 400)

cl <- pr(q10, ca); cm <- pr(q50, ca); ch <- pr(q90, ca)
cover <- function(s) mean(ca$y >= cm - s * (cm - cl) & ca$y <= cm + s * (ch - cm))
grid <- seq(0.5, 4, length.out = 400)
SCALE <- grid[which.min(abs(sapply(grid, cover) - 0.80))]

tl <- pr(q10, te); tm <- pr(q50, te); th <- pr(q90, te)
lo <- tm - SCALE * (tm - tl); hi <- tm + SCALE * (th - tm)
cat(sprintf("conformal scale %.2f | realised coverage of the nominal 80%% band: %.1f%%\n",
            SCALE, 100 * mean(te$y >= lo & te$y <= hi)))
cat(sprintf("median 80%% width %.4f of win rate\n", median(hi - lo)))
```
