---
title: "05 -- The three abstract figures"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE,
                      fig.path = "../figures/")
library(tidyverse); library(plotly); library(patchwork)
DATA <- "../data"; GAMMA <- 1.83; UR <- 0.31
wp <- function(x, ref) { num <- (ref + UR)^GAMMA; num / (num + pmax(x + UR, 1e-6)^GAMMA) }
stints  <- read_csv(file.path(DATA, "pitcher_stints_full.csv"),  show_col_types = FALSE)
seasons <- read_csv(file.path(DATA, "pitcher_seasons_full.csv"), show_col_types = FALSE)
FSTAR26 <- seasons %>% filter(season == 2026) %>% pull(FSTAR) %>% first()
ROLE_C <- c(Starter = "#2E6FB7", Relief = "#E8A33D", Closer = "#2E9E6B")
```

The abstract figures in the repository were produced by `scripts/bare_figs.py`, which uses
matplotlib's 3-D surfaces. R equivalents are given here: `plotly` for the two surfaces and
`ggplot2` for the gradient panel. The matplotlib versions remain canonical for the PDF,
because they are what the typeset abstract includes.

## Panel A -- the FIP_Pyth_WL output surface

```{r panelA}
d <- stints %>% filter(season == 2026, !is.na(FIP), IP >= 20) %>%
  mutate(G_opp = IP / 9, wr = wp(FIP, FSTAR26), fw = G_opp * wr)

ipg <- seq(15, 200, length.out = 80); fpg <- seq(1.0, 6.8, length.out = 80)
grid <- outer(ipg, fpg, function(i, f) (i / 9) * wp(f, FSTAR26))
grid[grid > 13] <- NA

plot_ly() %>%
  add_surface(x = ipg, y = fpg, z = t(grid), opacity = 0.55, colorscale = "Viridis",
              colorbar = list(title = "FIP-W")) %>%
  add_markers(data = d, x = ~IP, y = ~FIP, z = ~fw, color = ~role, colors = ROLE_C,
              marker = list(size = 3)) %>%
  layout(scene = list(xaxis = list(title = "Innings pitched (opportunity)"),
                      yaxis = list(title = "FIP (quality)"),
                      zaxis = list(title = "FIP-W (wins produced)")))
```

The dashed break-even ridge in the published figure is the plane FIP = F*, where a pitcher
is exactly .500.

## Panel B -- staff output per dollar across roles

```{r panelB}
PAR <- read_csv(file.path(DATA, "repr_params_full.csv"), show_col_types = FALSE)
MIN_SAL <- 0.78
SAL <- c(Starter = 6.80, Relief = 1.60, Closer = 8.00); SLOTS <- c(Starter = 5, Relief = 7, Closer = 1)
qf <- function(r, s) { a <- PAR %>% filter(role == r)
  a$w_max - (a$w_max - a$w_min) * exp(-a$k * (pmax(s, MIN_SAL) - MIN_SAL)) }
BUD <- sum(SAL * SLOTS)
staff <- function(rt, ct) {
  st <- BUD - rt - ct
  ifelse(st < 5 * MIN_SAL | rt < 7 * MIN_SAL | ct < MIN_SAL, NA,
         5 * qf("Starter", st / 5) + 7 * qf("Relief", rt / 7) + qf("Closer", ct))
}
rg <- seq(7 * MIN_SAL, 36, length.out = 90); cg <- seq(MIN_SAL, 15, length.out = 90)
z <- outer(rg, cg, Vectorize(staff))
plot_ly(x = rg, y = cg, z = t(z)) %>% add_surface(colorscale = "Viridis", opacity = 0.65) %>%
  layout(scene = list(xaxis = list(title = "Bullpen $M"),
                      yaxis = list(title = "Closer $M"),
                      zaxis = list(title = "Staff FIP-W")))
```

## Figure 2 -- what a dollar buys, and the price of a win-flip

```{r fig2, fig.width=9, fig.height=3}
sal <- seq(MIN_SAL, 11, length.out = 400)
dq <- function(r, s) { a <- PAR %>% filter(role == r)
  a$k * (a$w_max - a$w_min) * exp(-a$k * (pmax(s, MIN_SAL) - MIN_SAL)) }
g <- expand_grid(role = PAR$role, s = sal) %>% rowwise() %>%
  mutate(w = qf(role, s), flip = 1 / dq(role, s)) %>% ungroup()

p1 <- ggplot(g, aes(s, w, colour = role)) + geom_line(linewidth = 1) +
  labs(x = "Salary per roster place ($M)", y = "FIP-W the place produces") +
  theme_minimal(base_size = 9) + theme(legend.position = "bottom")
p2 <- ggplot(filter(g, flip <= 13), aes(s, flip, colour = role)) + geom_line(linewidth = 1) +
  geom_hline(yintercept = 3.56, linetype = "dashed") +
  labs(x = "Salary per roster place ($M)", y = "$M per win-flip") +
  theme_minimal(base_size = 9) + theme(legend.position = "bottom")
p1 + p2
```
